---
title: "If/Else"
format: html
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
```

# If/Else Statements

When we are writing loops or function, it can be useful to take different actions in response to different values or situations. That's where if/else statements come in.

You may be familiar with the `ifelse()` function, which can work with vectors:

```{r}
ages <- c(16, 17, 18, 19, 20, 21, 22, 23)

# ifelse(conditional test, value if true, value if false)

ifelse(ages >= 18, "adult", "child")
```

When we're working in a loop or a function, we may want to do more than just return one of two different values.

```{r, eval=FALSE}
if (conditional test) {
  do_something()
} else {
  do_something_else()
}
```

Example:

```{r}
x <- sample.int(10, 1) # random integer 1-10
print(paste("x is", x))

if (x >= 9) {
  print("Winner!")
} else {
  print("Try again")
}
```

Note that the condition with `if()` evaluates to a **single** TRUE or FALSE value, not a vector of values.

The `else()` part is optional:

```{r}
x <- sample.int(10, 1) # random integer 1-10
print(paste("x is", x))

if (x >= 9) {
  print("Winner!")
}
```

## TRY IT

Write a function that returns the square root of a number. Check the value of the input, and if it's negative, print "That's not real!". Only return an answer if the supplied value is \>= 0.

```{r}
safe_sqrt = function(x) {
  if (x < 0) {
    print("That's not real!")
  } else {
    return(sqrt(x))
  }
}

safe_sqrt(25)
safe_sqrt(-1)
```

## TRY IT

Write a for loop to print a welcome message to each name in a vector. If you encounter a missing value, ignore it (don't print a message).

Note: the command `next` ends the current iteration of the loop and moves on to the next value. You don't have to use it to accomplish this, but it's an option.

```{r}
people <- c("Aisha", "Daniel", "Priya", NA, "Miguel", "Luca")

# potentially helpful code
cat("Hi there", "Joe", "\n")  

for (name in people) {
  if (is.na(name)) {
    next # skip this iteration
  }
  cat("Hi there", name, "\n")
}
```

Challenge: create several different welcome/hello messages and randomly select one to use with each name. Look at the `sample()` function to help you with this.

```{r}
people <- c("Aisha", "Daniel", "Priya", NA, "Miguel", "Luca")

greetings = c("Hi there", "Hello", "Hey", "welcome")

# potentially helpful code
cat("Hi there", "Joe", "\n")  

for (name in people) {
  if (is.na(name)) {
    next # skip this iteration
  }
  greeting = sample(greetings, 1)
  cat(greeting, name, "\n")
}
```

## Multiple Conditions

If/else statements can be combined to check multiple conditions:

```{r}
x <- sample.int(10, 1)
print(paste("x is", x))


if (x == 10) {
  print("You win the grand prize!")  ## only for x == 10
} else if (x >= 8) {
  print("You win the runner up prize.")  ## for x == 9 or x == 8
} else {
  print("Sorry, you lose.")
}
```

For each value, only one of these print statements can execute. So while 10 is \>= 8, if x is 10, the first print statement will execute, and the "else" to that will not.

## TRY IT

Write a function that returns a letter grade based on the percent (0-100) score provided as input.

```{r}
letter_grade = function(score) {
  if (score >= 90) {
    return("A")
  } else if (score >= 80) {
    return("B")
  } else if (score >= 70) {
    return("C")
  } else if (score >= 60) {
    return("D")
  } else {
    return("F")
  }
}

letter_grade(95)
letter_grade(82)
letter_grade(71)
letter_grade(55)
```
